\pkg{cv}: An \proglang{R} Package for Cross-Validation of Regression Models

### 2024-01-10 scratch/notes_john/JSS-article/JSS-article-reduced.Rmd
edit JSS paper

---
documentclass: jss
author:
  - name: John Fox
    orcid: 0000-0002-1196-8012
    affiliation: 'McMaster University'
    address: |
      | Hamilton, Ontario, Canada
    email: \email{jfox@mcmaster.ca}
    url: https://www.john-fox.ca/
  - name: Georges Monette
    orcid: 0000-0000-0000-0000
    affiliation: 'York University'
    # To add another line, use \AND at the end of the previous one as above
    address: |
      | Toronto, Ontario, Canada
title:
  formatted: "\\pkg{cv}: An \\proglang{R} Package for Cross-Validation of Regression Models"
  # If you use tex in the formatted title, also supply version without
  plain:     "cv: An R Package for Cross-Validation of Regression Models"
  # For running headers, if needed
  short:     "\\pkg{cv}: Cross-Validation"
abstract: >
  The abstract of the article.
keywords:
  # at least one keyword must be supplied
  formatted: [cross-validation, regression analysis, model selection, "\\proglang{R}"]
  plain:     [cross-validation, regression analysis, model selection, R]
preamble: >
  \usepackage{amsmath}
output: 
  rticles::jss_article:
    extra_dependencies: "subfig"
bibliography: ["cv.bib"]
---

```{r, setup, include=FALSE}
options(prompt = 'R> ', continue = '+ ')
```

# Cross-validation

@@

Validating research by replication on independently collected data is a common scientific norm. Emulating this process in a single study by data-division is less common: The data are randomly divided into two, possibly equal-size, parts; the first part is used to develop and fit a statistical model; and then the second part is used to assess the adequacy of the model fit to the first part of the data. Data-division, however, suffers from two problems: (1) Dividing the data decreases the sample size and thus increases sampling error; and (2), even more disconcertingly, particularly in smaller samples, the results can vary substantially based on the random division of the data: See @Harrell:2015 [Sec. 5.3] for this and other remarks about data-division and cross-validation.

Cross-validation speaks to both of these issues. In CV, the data are randomly divided as equally as possible into several, say $k$, parts, called "folds." The statistical model is fit $k$ times, leaving each fold out in turn. Each fitted model is then used to predict the response variable for the cases in the omitted fold. A CV criterion or "cost" measure, such as the mean-squared error ("MSE") of prediction, is then computed using these predicted values. In the extreme $k = n$, the number of cases in the data, thus omitting individual cases and refitting the model $n$ times---a procedure termed "leave-one-out (LOO) cross-validation."

Because the $n$ models are each fit to $n - 1$ cases, LOO CV produces a nearly unbiased estimate of prediction error. The $n$ regression models are highly statistical dependent, however, based as they are on nearly the same data, and so the resulting estimate of prediction error has relatively large variance. In contrast, estimated prediction error for $k$-fold CV with $k = 5$ or $10$ (commonly employed choices) are somewhat biased but have smaller variance. It is also possible to correct $k$-fold CV for bias (see below).

# Examples

## Polynomial regression for the Auto data

The data for this example are drawn from the \pkg{ISLR2} package for \proglang{R}, associated with @JamesEtAl:2021. The presentation here is close (though not identical) to that in the original source [@JamesEtAl:2021 Secs. 5.1, 5.3], and it demonstrates the use of the `cv()` function in the \pkg{cv} package.[^boot]

@@

The `Auto` dataset contains information about 392 cars:

```{r Auto}
data("Auto", package="ISLR2")
head(Auto)
dim(Auto)
```
With the exception of `origin` (which we don't use here), these variables are largely self-explanatory, except possibly for units of measurement: for details see `help("Auto", package="ISLR2")`.

We'll focus here on the relationship of `mpg` (miles per gallon) to `horsepower`, as displayed in the following scatterplot:

The relationship between the two variables is monotone, decreasing, and nonlinear. Following @JamesEtAl:2021, we'll consider approximating the relationship by a polynomial regression, with the degree of the polynomial $p$ ranging from 1 (a linear regression) to 10.[^log-trans] Polynomial fits  for $p = 1$ to $5$ are shown in the following figure:

[^log-trans]: Although it serves to illustrate the use of CV, a polynomial is probably not the best choice here. Consider, for example the scatterplot for log-transformed `mpg` and `horsepower`, produced by `plot(mpg ~ horsepower, data=Auto, log="xy")` (execution of which is left to the reader).

```{r mpg-horsepower-scatterplot-polynomials}
#| out.width = "50%",
#| fig.height = 5,
#| fig.cap = "`mpg` vs `horsepower` for the `Auto` data"
plot(mpg ~ horsepower, data=Auto)
horsepower <- with(Auto, 
                   seq(min(horsepower), max(horsepower), 
                       length=1000))
for (p in 1:5){
  m <- lm(mpg ~ poly(horsepower,p), data=Auto)
  mpg <- predict(m, newdata=data.frame(horsepower=horsepower))
  lines(horsepower, mpg, col=p + 1, lty=p, lwd=2)
}
legend("topright", legend=1:5, col=2:6, lty=1:5, lwd=2,
       title="Degree", inset=0.02)
```
The linear fit is clearly inappropriate; the fits for $p = 2$ (quadratic) through $4$ are very similar; and the fit for $p = 5$ may over-fit the data by chasing one or two relatively high `mpg` values at the right (but see the CV results reported below).

The following graph shows two measures of estimated (squared) error as a function of polynomial-regression degree: The mean-squared error ("MSE"), defined as $\mathsf{MSE} = \frac{1}{n}\sum_{i=1}^n (y_i - \widehat{y}_i)^2$, and the usual residual variance, defined as $\widehat{\sigma}^2 = \frac{1}{n - p - 1} \sum_{i=1}^n (y_i - \widehat{y}_i)^2$. The former necessarily declines with $p$ (or, more strictly, can't increase with $p$), while the latter gets slightly larger for the largest values of $p$, with the "best" value, by a small margin, for $p = 7$.

```{r mpg-horsepower-MSE-se}
#| out.width = "50%",
#| fig.height = 5,
#| fig.cap = "Estimated squared error as a function of polynomial degree, $p$"
library("cv") # for mse() and other functions

@@

plot(c(1, 10), range(mse, var), type="n",
     xlab="Degree of polynomial, p",
     ylab="Estimated Squared Error")
lines(1:10, mse, lwd=2, lty=1, col=2, pch=16, type="b")
lines(1:10, var, lwd=2, lty=2, col=3, pch=17, type="b")
legend("topright", inset=0.02,
       legend=c(expression(hat(sigma)^2), "MSE"),
       lwd=2, lty=2:1, col=3:2, pch=17:16)
```
The code for this graph uses the `mse()` function from the \pkg{cv} package to compute the MSE for each fit.

### Using cv()

The generic `cv()` function has an `"lm"` method, which by default performs $k = 10$-fold CV:
```{r cv-lm-1}
m.auto <- lm(mpg ~ poly(horsepower, 2), data=Auto)
summary(m.auto)

cv(m.auto)
```
The \code{"lm"} method by default uses \code{mse()} as the CV criterion and the Woodbury matrix identity to update the regression with each fold deleted without having literally to refit the model. Computational details are discussed in the final section of this vignette. The function reports the CV estimate of MSE, a biased-adjusted estimate of the MSE (the bias adjustment is explained in the final section), and the MSE is also computed for the original, full-sample regression. Because the division of the data into 10 folds is random, \code{cv()} explicitly (randomly) generates and saves a seed for \proglang{R}'s pseudo-random number generator, to make the results replicable. The user can also specify the seed directly via the \code{seed} argument to \code{cv()}.

To perform LOO CV, we can set the \code{k} argument to \code{cv()} to the number of cases in the data, here \code{k=392}, or, more conveniently, to \code{k="loo"} or \code{k="n"}:

```{r cv.lm-2`}
cv(m.auto, k="loo")
```
For LOO CV of a linear model, \code{cv()} by default uses the hatvalues from the model fit to the full data for the LOO updates, and reports only the CV estimate of MSE. Alternative methods are to use the Woodbury matrix identity or the "naive" approach of literally refitting the model with each case omitted. All three methods produce exact results for a linear model (within the precision of floating-point computations):
```{r cv.lm-3}
cv(m.auto, k="loo", method="naive")

cv(m.auto, k="loo", method="Woodbury")
```
The \code{"naive"} and \code{"Woodbury"} methods also return the bias-adjusted estimate of MSE and the full-sample MSE, but bias isn't an issue for LOO CV.

This is a small regression problem and all three computational approaches are essentially instantaneous, but it is still of interest to investigate their relative speed. In this comparison, we include the \code{cv.glm()} function from the \pkg{boot} package, which takes the naive approach, and for which we have to fit the linear model as an equivalent Gaussian GLM. We use the \code{microbenchmark()} function from the package of the same name for the timings [@Mersmann:2023]:
```{r cv.lm.timings, cache=TRUE}
m.auto.glm <- glm(mpg ~ poly(horsepower, 2), data=Auto)
boot::cv.glm(Auto, m.auto.glm)$delta

@@

[^microbenchmark]: Out of impatience, we asked \code{microbenchmark()} to execute each command only 10 times rather than the default 100. With the exception of the last columns, the output is self-explanatory. The last column shows which methods have average timings that are statistically distinguishable. Because of the small number of repetitions (i.e., 10), the \code{"hatvalues"} and \code{"Woodbury"} methods aren't distinguishable, but the difference between these methods persists when we perform more repetitions---we invite the reader to redo this computation with the default \code{times=100} repetitions.

### Comparing competing models

@@

We illustrate with the polynomial regression models of varying degree for the \code{Auto} data (discussed previously), beginning by fitting and saving the 10 models:
```{r polyomial-models}
for (p in 1:10){
  assign(paste0("m.", p),
         lm(mpg ~ poly(horsepower, p), data=Auto))
}
objects(pattern="m\\.[0-9]")
summary(m.2) # for example, the quadratic fit
```
We then apply \code{cv()} to the list of 10 models (the \code{data} argument is required):
```{r polynomial-regression-CV}
# 10-fold CV
cv.auto.10 <- cv(models(m.1, m.2, m.3, m.4, m.5,
                     m.6, m.7, m.8, m.9, m.10),
              data=Auto, seed=2120)
cv.auto.10[1:2] # for the linear and quadratic models

# LOO CV
cv.auto.loo <- cv(models(m.1, m.2, m.3, m.4, m.5,
                        m.6, m.7, m.8, m.9, m.10),
                 data=Auto, k="loo")
cv.auto.loo[1:2] # linear and quadratic models
```
Because we didn't supply names for the models in the calls to the \code{models()} function, the names \code{model.1}, \code{model.2}, etc., are generated by the function.

Alternatively, we can use the \code{plot()} method for \code{"cvModList"} objects to compare the models, though with separate graphs for 10-fold and LOO CV:

```{r, polynomial-regression-CV-graph-2, fig.show="hold"}
#| fig.height = 5,
#| fig.cap = "Cross-validated (a) 10-fold and (b) LOO MSE as a function of polynomial degree, $p$",
#| fig.subcap=c('', ''),
#| fig.ncol = 2, out.width = "50%", fig.align = "center"
plot(cv.auto.10, main="Polynomial Regressions, 10-Fold CV",
     axis.args=list(labels=1:10), xlab="Degree of Polynomial, p")
plot(cv.auto.loo, main="Polynomial Regressions, LOO CV",
     axis.args=list(labels=1:10), xlab="Degree of Polynomial, p")
```

In this example, 10-fold and LOO CV produce generally similar results, and also results that are similar to those produced by the estimated error variance $\widehat{\sigma}^2$ for each model, reported above (except for the highest-degree polynomials, where the CV results more clearly suggest over-fitting).

@@

The fundamental analogy for cross-validation is to the collection of new data. That is, predicting the response in each fold from the model fit to data in the other folds is like using the model fit to all of the data to predict the response for new cases from the values of the predictors for those new cases. As we explained, the application of this idea to independently sampled cases is straightforward---simply partition the data into random folds of equal size and leave each fold out in turn, or, in the case of LOO CV, simply omit each case in turn.

In contrast, mixed-effects models are fit to *dependent* data, in which cases as clustered, such as hierarchical data, where the clusters comprise higher-level units (e.g., students clustered in schools), or longitudinal data, where the clusters are individuals and the cases repeated observations on the individuals over time.[^crossed-effects]

[^crossed-effects]: There are, however, more complex situations that give rise to so-called *crossed* (rather than *nested*) random effects. For example, consider students within classes within schools. In primary schools, students typically are in a single class, and so classes are nested within schools. In secondary schools, however, students typically take several classes and students who are together in a particular class may not be together in other classes; consequently, random effects based on classes within schools are crossed. The \code{lmer()} function in the \pkg{lme4} package is capable of modeling both nested and crossed random effects, and the \code{cv()} methods for mixed models in the \pkg{cv} package pertain to both nested and crossed random effects. We present an example of the latter later in the vignette.

@@

2. Treat CV as analogous to predicting the response for a newly observed case in an *existing cluster*. In this instance, the folds comprise one or more individual cases, and the predictions can use both the fixed and random effects.

@@

The HSB data are included in the \code{MathAchieve} and \code{MathAchSchool} data sets in the \pkg{nlme} package  [@PinheiroBates:2000]. \code{MathAchieve} includes individual-level data on 7185  students in 160 high schools, and \code{MathAchSchool} includes school-level data:
```{r HSB-data}
data("MathAchieve", package="nlme")
dim(MathAchieve)
head(MathAchieve, 3)
tail(MathAchieve, 3)

@@

Some data-management is required before fitting a mixed-effects model to the HSB data, for which we use the \pkg{dplyr} package [@WickhamEtAl:2023]:
```{r HSB-data-management, cache=TRUE}
library("dplyr")
MathAchieve %>% group_by(School) %>%
  summarize(mean.ses = mean(SES)) -> Temp
Temp <- merge(MathAchSchool, Temp, by="School")
HSB <- merge(Temp[, c("School", "Sector", "mean.ses")],
             MathAchieve[, c("School", "SES", "MathAch")], by="School")
names(HSB) <- tolower(names(HSB))

HSB$cses <- with(HSB, ses - mean.ses)
```
In the process, we created two new school-level variables: \code{meanses}, which is the average SES for students in each school; and \code{cses}, which is school-average SES centered at its mean. For details, see @FoxWeisberg:2019 [Sec. 7.2.2].

Still following Fox and Weisberg, we proceed to use the \code{lmer()} function in the \pkg{lme4} package [@BatesEtAl:2015] to fit a mixed model for math achievement to the HSB data:
```{r HSB-lmer, cache=TRUE}
library("lme4")
hsb.lmer <- lmer(mathach ~ mean.ses*cses + sector*cses
                   + (cses | school), data=HSB)
summary(hsb.lmer, correlation=FALSE)
```

@@

If the \code{clusterVariables} argument is omitted, then case-level CV is employed, with \code{k = 10} folds as the default, here each with $7185/10 \approx 719$ students. Notice that one of the 10 models refit with a fold removed failed to converge. Convergence problems are common in mixed-effects modeling. The apparent issue here is that an estimated variance component is close to or equal to 0, which is at a boundary of the parameter space. That shouldn't disqualify the fitted model for the kind of prediction required for cross-validation.

There is also a \code{cv()} method for linear mixed models fit by the \code{lme()} function in the \pkg{nlme} package, and the arguments for \code{cv()} in this case are the same as for a model fit by \code{lmer()} or \code{glmer()}. We illustrate with the mixed model fit to the HSB data:
```{r hsb-lme, cache=TRUE}
library(nlme)
hsb.lme <- lme(mathach ~ mean.ses*cses + sector*cses,
                 random = ~ cses | school, data=HSB,
               control=list(opt="optim"))
summary(hsb.lme)

cv(hsb.lme, k=10, clusterVariables="school", seed=5240)

cv(hsb.lme, seed=1575)
```
We used the same random-number generator seeds as in the previous example cross-validating the model fit by \code{lmer()}, and so the same folds are employed in both cases.[^optimizer] The estimated covariance components and fixed effects in the summary output differ slightly between the \code{lmer()} and \code{lme()} solutions, although both functions seek to maximize the REML criterion. This is, of course, to be expected when different algorithms are used for numerical optimization.  To the precision reported, the cluster-level CV results for the \code{lmer()} and \code{lme()} models are identical, while the case-level CV results are very similar but not identical.

[^optimizer]: The observant reader will notice that we set the argument \code{control=list(opt="optim")} in the call to \code{lme()}, changing the optimizer employed from the default \code{"nlminb"}. We did this because with the default optimizer, \code{lme()} encountered the same convergence issue as \code{lmer()}, but rather than issuing a warning, \code{lme()} failed, reporting an error. As it turns out, setting the optimizer to \code{"optim"} avoids this problem.

@@

We introduce an artificial data set that exemplifies aspects of cross-validation particular to hierarchical models. Using this data set, we show that model comparisons employing cluster-based and those employing case-based cross-validation may not agree on a "best" model. Furthermore, commonly used measures of fit, such as mean-squared error, do not necessarily become smaller as models become larger, even when the models are nested, and even when the measure of fit is computed for the whole data set.

>>>>

---
documentclass: jss
author:
  - name: John Fox
    orcid: 0000-0002-1196-8012
    affiliation: 'McMaster University'
    address: |
      | Hamilton, Ontario, Canada
    email: \email{jfox@mcmaster.ca}
    url: https://www.john-fox.ca/
  - name: Georges Monette
    orcid: 0000-0000-0000-0000
    affiliation: 'York University'
    # To add another line, use \AND at the end of the previous one as above
    address: |
      | Toronto, Ontario, Canada
title:
  formatted: "\\pkg{cv}: An \\proglang{R} Package for Cross-Validation of Regression Models"
  # If you use tex in the formatted title, also supply version without
  plain:     "cv: An R Package for Cross-Validation of Regression Models"
  # For running headers, if needed
  short:     "\\pkg{cv}: Cross-Validation of Regression Models"
abstract: >
  The abstract of the article.
keywords:
  # at least one keyword must be supplied
  formatted: [cross-validation, regression analysis, model selection, "\\proglang{R}"]
  plain:     [cross-validation, regression analysis, model selection, R]
preamble: >
  \usepackage{amsmath}
output: 
  rticles::jss_article:
    extra_dependencies: "subfig"
bibliography: ["cv.bib"]
---

```{r, setup, include=FALSE}
options(prompt = 'R> ', continue = '+ ')
palette(car::carPalette())
```

# Introduction

@@

Validating research by replication on independently collected data is a common scientific norm. Emulating this process in a single study by data-division is less common: The data are randomly divided into two, possibly equal-size, parts; the first part is used to develop and fit a statistical model; and then the second part is used to assess the adequacy of the model fit to the first part of the data. Data-division, however, suffers from two problems: (1) Dividing the data decreases the sample size and thus increases sampling error; and (2), even more disconcertingly, the results can vary substantially based on the random division of the data, particularly in smaller samples: See @Harrell:2015 [Sec. 5.3] for this and other remarks about data-division and cross-validation.

Cross-validation speaks to both of these issues. In CV, the data are randomly divided as equally as possible into several, say $k$, parts, called "folds." The statistical model is fit $k$ times, leaving each fold out in turn. Each fitted model is then used to predict the response variable for the cases in the omitted fold. A CV criterion (also termed a "cost" or "loss" measure), such as the mean-squared error ("MSE") of prediction, is then computed using these predicted values. In the extreme $k = n$, the number of cases in the data, thus omitting individual cases and refitting the model $n$ times---a procedure termed "leave-one-out (LOO) cross-validation."

Because the $n$ models are each fit to $n - 1$ cases, LOO CV produces a nearly unbiased estimate of prediction error. The $n$ regression models are highly statistical dependent, however, based as they are on nearly the same data, and so the resulting estimate of prediction error has relatively large variance. In contrast, estimated prediction error for $k$-fold CV with $k = 5$ or $10$ (commonly employed choices) are somewhat biased but have smaller variance. It is also possible to correct $k$-fold CV for bias (see Section \ref{computation-of-the-bias-corrected-cv-criterion-and-confidence-intervals}).

The \pkg{cv} package for \proglang{R} automates the process of cross-validation for standard \proglang{R} statistical model objects. The principal function in the packages, also named \code{cv()}, has methods for objects produced by a number of commonly employed regression-modeling functions, including those for mixed-effects models:
```{r cv-methods}
library("cv", quietly=TRUE)
methods("cv")
```
The \code{"modList"} method for \code{cv()} cross-validates several competing models, not necessarily of the same class, using the same division of the data into folds. The \code{cv()} function is introduced in the context of a preliminary example in Section \ref{preliminary-example-polynomial-regression} of the paper.

Cross-validating mixed-effects models involves special considerations that we take up in Section \ref{cross-validating-mixed-effects-models}.

The \code{cvSelect()} function, discussed in Section \ref{cross-validating-model-selection}, cross-validates a complex model-specification process that may, for example, involve choice of data transformations and predictors.

The \code{"default"} \code{cv()} method works (perhaps with a bit of coaxing) with many other existing regression-model classes for which there is an \code{update()} method that accepts a \code{data} argument. More generally, the \pkg{cv} package is designed to be extensible, as discussed in Section \ref{extending-the-cv-package}.

In the interest of brevity, we won't describe all of the features of the \pkg{cv} package here, concentrating on the aspects of the package that are relatively novel. For example, the \code{cv()} and \code{cvSelect()} functions can perform computations in parallel and can independently replicate a cross-validation procedure several times. These and other features not discussed in this paper are taken up in the vignettes distributed with the package, which also provides greater detail on some of topics that we do describe, such as extensions to the package.

# Preliminary Example: Polynomial regression

The data for the example in this section are drawn from the \pkg{ISLR2} package for \proglang{R}, associated with @JamesEtAl:2021. The presentation here is close (though not identical) to that in the original source [@JamesEtAl:2021 Secs. 5.1, 5.3], and it demonstrates the use of the `cv()` function.[^boot]

@@

The \code{Auto} dataset contains information about 392 cars:

```{r Auto}
data("Auto", package="ISLR2")
summary(Auto)
```
With the exception of `origin` (which we don't use here), these variables are largely self-explanatory, except possibly for units of measurement: for details see `help("Auto", package="ISLR2")`.

```{r mpg-horsepower-scatterplot-polynomials, echo=FALSE}
#| out.width = "50%",
#| fig.height = 5,
#| fig.cap = "mpg vs horsepower for the Auto data, showing fitted polynomials of degree 1 through 5."
plot(mpg ~ horsepower, data=Auto)
horsepower <- with(Auto, 
                   seq(min(horsepower), max(horsepower), 
                       length=1000))
for (p in 1:5){
  m <- lm(mpg ~ poly(horsepower,p), data=Auto)
  mpg <- predict(m, newdata=data.frame(horsepower=horsepower))
  lines(horsepower, mpg, col=p + 1, lty=p, lwd=2)
}
legend("topright", legend=1:5, col=2:6, lty=1:5, lwd=2,
       title="Degree", inset=0.02)
```

We'll focus here on the relationship of \code{mpg} (miles per gallon) to \code{horsepower}, as displayed in Figure \ref{fig:mpg-horsepower-scatterplot-polynomials}. The relationship between the two variables is monotone, decreasing, and nonlinear. Following @JamesEtAl:2021, we'll consider approximating the relationship by a polynomial regression, with the degree of the polynomial $p$ ranging from 1 (a linear regression) to 10.[^log-trans] Polynomial fits  for $p = 1$ to $5$ are shown in Figure \ref{fig:mpg-horsepower-scatterplot-polynomials}. The linear fit is clearly inappropriate; the fits for $p = 2$ (quadratic) through $4$ are very similar; and the fit for $p = 5$ may over-fit the data by chasing one or two relatively high `mpg` values at the right (but see the CV results reported below).

[^log-trans]: Although it serves to illustrate the use of CV, a polynomial is not the best choice here. Consider, for example the scatterplot for log-transformed `mpg` and `horsepower`, produced by `plot(mpg ~ horsepower, data=Auto, log="xy")` (execution of which is left to the reader). We revisit the \code{Auto} data in Section \ref{cross-validating-model-selection}.

Figure \ref{fig:mpg-horsepower-MSE-se} shows two measures of estimated (squared) error as a function of polynomial-regression degree: The mean-squared error ("MSE"), defined as $\mathsf{MSE} = \frac{1}{n}\sum_{i=1}^n (y_i - \widehat{y}_i)^2$, and the usual residual variance, defined as $\widehat{\sigma}^2 = \frac{1}{n - p - 1} \sum_{i=1}^n (y_i - \widehat{y}_i)^2$. The former necessarily declines with $p$ (or, more strictly, can't increase with $p$), while the latter gets slightly larger for the largest values of $p$, with the "best" value, by a small margin, for $p = 7$.

```{r mpg-horsepower-MSE-se, echo=FALSE}
#| out.width = "50%",
#| fig.height = 5,
#| fig.cap = "Estimated squared error as a function of polynomial degree, $p$"

@@

plot(c(1, 10), range(mse, var), type="n",
     xlab="Degree of polynomial, p",
     ylab="Estimated Squared Error")
lines(1:10, mse, lwd=2, lty=1, col=2, pch=16, type="b")
lines(1:10, var, lwd=2, lty=2, col=3, pch=17, type="b")
legend("topright", inset=0.02,
       legend=c(expression(hat(sigma)^2), "MSE"),
       lwd=2, lty=2:1, col=3:2, pch=17:16)
```

The generic `cv()` function has an `"lm"` method,
```{r cv-lm-method}
args(cv:::cv.lm)
```
which takes the following arguments:

* \code{model}, an \code{"lm"} object, the only required argument.

* \code{data}, which can usually be inferred from the \code{model} object.

* \code{criterion}, a function to compute the CV criterion (defaulting to \code{mse}).

* \code{k}, the number of folds to employ (defaulting to \code{10}); the character value \code{"n"} or \code{"loo"} may be supplied to specify leave-one-out cross-validation.

* \code{reps}, the number of times to repeat the CV procedure (defaulting to 1).

* \code{seed}, the seed for \proglang{R}'s pseudo-random number generator; if not specified a value is randomly selected, reported, and saved, so that the CV procedure is replicable.

* \code{confint}, whether or not to compute a confidence interval for the CV criterion, defaulting to \code{TRUE} if there are at least 400 cases; a confidence interval is computed only if the CV criterion can be expressed as the average of casewise components (see Section \ref{computation-of-the-bias-corrected-cv-criterion-and-confidence-intervals} for details).

* \code{level}, the level for the confidence interval (defaulting to \code{0.95}).

* \code{method}, the computational method to employ: \code{"hatvalues"} is relevant only for LOO CV and bases computation on the hatvalues for the linear model; \code{"Woodbury"} employs the Woodbury matrix to compute the CV criterion with each fold deleted; \code{"naive"} updates the model using the \code{update()} function; and \code{"auto"} selects \code{"hatvalues"} for LOO CV and \code{"Woodbury"} for $k$-fold CV, both of which are much more efficient than literally updating the least-squares fit (see below and Section \ref{efficient-computation-for-linear-and-generalized-linear-models}).

* \code{ncores}, the number of cores to employ for parallel computation; if \code{cores = 1} (the default), the computations are not parallelized.

To illustrate, we perform 10-fold CV for a quadratic polynomial fit to the \code{Auto} data:
```{r cv-lm-1}
m.auto <- lm(mpg ~ poly(horsepower, 2), data=Auto)
summary(m.auto)

cv(m.auto, confint=TRUE)
```
The function reports the CV estimate of MSE, a bias-adjusted estimate of the MSE (the bias adjustment is explained in Section \ref{computation-of-the-bias-corrected-cv-criterion-and-confidence-intervals}), and the MSE is also computed for the original, full-sample regression. Because the number of cases $n = 392 < 400$ for the \code{Auto} data, we set the argument \code{confint=TRUE} to obtain a confidence interval for the MSE, which proves to be quite wide.

To perform LOO CV:

```{r cv.lm-2`}
cv(m.auto, k="loo")
```
The \code{"hatvalues"} method reports only the CV estimate of MSE. Alternative methods are to use the Woodbury matrix identity or the "naive" approach of literally refitting the model with each case omitted. All three methods produce exact results for a linear model (within the precision of floating-point computations):
```{r cv.lm-3}
cv(m.auto, k="loo", method="naive", confint=TRUE)

cv(m.auto, k="loo", method="Woodbury", confint=TRUE)
```
The \code{"naive"} and \code{"Woodbury"} methods also return the bias-adjusted estimate of MSE (and a confidence interval around it) along with the full-sample MSE, but bias isn't an issue for LOO CV.

This is a small regression problem and all three computational approaches are essentially instantaneous, but it is still of interest to investigate their relative speed. In the following comparison, we include the \code{cv.glm()} function from the \pkg{boot} package, which takes the naive approach, and for which we have to fit the linear model as an equivalent Gaussian GLM. We use the \code{microbenchmark()} function from the package of the same name  [@Mersmann:2023] for the timings:
```{r cv.lm.timings, cache=TRUE}
m.auto.glm <- glm(mpg ~ poly(horsepower, 2), data=Auto)
boot::cv.glm(Auto, m.auto.glm)$delta

@@

[^microbenchmark]: Out of impatience, we asked \code{microbenchmark()} to execute each command only 10 times rather than the default 100. With the exception of the last column, the output is self-explanatory. The last column shows which methods have average timings that are statistically distinguishable. Because of the small number of repetitions (i.e., 10), the \code{"hatvalues"} and \code{"Woodbury"} methods aren't distinguishable, but the difference between these methods persists when we perform more repetitions---we invite the reader to redo this computation with the default \code{times=100} repetitions.

## Comparing competing models

@@

We illustrate with the polynomial regression models of varying degree for the \code{Auto} data, beginning by fitting and saving the 10 models:
```{r polyomial-models}
mlist <- vector(10, mode="list")
for (p in 1:10) mlist[[p]] <- lm(mpg ~ poly(horsepower, p), data = Auto)
names(mlist) <- paste0("m.", 1:10)
mlist[2] # e.g., the quadratic fit
```
We then apply \code{cv()} to the list of 10 models (the \code{data} argument is required):
```{r polynomial-regression-CV}
# 10-fold CV
mlist <- do.call(models, mlist) # create "modList" object

cv.auto.10 <- cv(mlist, data=Auto, seed=2120)
cv.auto.10[2] # e.g., for quadratic model

# LOO CV
cv.auto.loo <- cv(mlist, data=Auto, k="loo")
cv.auto.loo[2] # e.g., for quadratic model
```
The \code{models()} function takes an arbitrary number of regression models as its arguments, which are optionally named, to create a \code{"modList"} object. Because we generated the polynomial regression models in a named list, we conveniently employ \code{do.call()} to supply the models as arguments to \code{models()}. Note that the names created for the list (e.g., \code{"m.2"}) are then used for the models. We can also invoke the \code{plot()} method for \code{"cvModList"} objects to compare the models (see Figure \ref{fig:polynomial-regression-CV-graph-2}):

```{r, polynomial-regression-CV-graph-2, fig.show="hold"}
#| fig.height = 5,
#| fig.cap = "Cross-validated (a) 10-fold and (b) LOO MSE as a function of polynomial degree, $p$.",
#| fig.subcap=c('', ''),
#| fig.ncol = 2, out.width = "50%", fig.align = "center"
plot(cv.auto.10, main="Polynomial Regressions, 10-Fold CV",
     axis.args=list(labels=1:10), xlab="Degree of Polynomial, p")
plot(cv.auto.loo, main="Polynomial Regressions, LOO CV",
     axis.args=list(labels=1:10), xlab="Degree of Polynomial, p")
```

In this example, 10-fold and LOO CV produce generally similar results, and also results that are similar to those produced by the estimated error variance $\widehat{\sigma}^2$ for each model (cf., Figure \ref{fig:mpg-horsepower-MSE-se} on page \pageref{fig:mpg-horsepower-MSE-se}), except for the highest-degree polynomials, where the CV results more clearly suggest over-fitting.

@@

The fundamental analogy for cross-validation is to the collection of new data. That is, predicting the response in each fold from the model fit to data in the other folds is like using the model fit to all of the data to predict the response for new cases from the values of the predictors for those new cases. As we explained, the application of this idea to independently sampled cases is straightforward.

In contrast, mixed-effects models are fit to *dependent* data, in which cases are clustered, such as hierarchical data, where the clusters comprise higher-level units (e.g., students clustered in schools), or longitudinal data, where the clusters are individuals and the cases are repeated observations on the individuals over time.[^crossed-effects]

[^crossed-effects]: There are, however, more complex situations that give rise to so-called *crossed* (rather than *nested*) random effects. For example, consider students within classes within schools. In primary schools, students typically are in a single class, and so classes are nested within schools. In secondary schools, however, students typically take several classes and students who are together in a particular class may not be together in other classes; consequently, random effects based on classes within schools are crossed. The \code{lmer()} function in the \pkg{lme4} package is capable of modeling both nested and crossed random effects, and the \code{cv()} methods for mixed models in the \pkg{cv} package pertain to both nested and crossed random effects. We present an example of the latter in a vignette for the \pkg{cv} package.

@@

2. Treat CV as analogous to predicting the response for a newly observed case in an *existing cluster*. In this instance, the folds comprise one or more individual cases, and the predictions can use both the fixed and random effects---so-called "best-linear-unbiased predictors" or "BLUPs."

@@

The HSB data are included in the \code{MathAchieve} and \code{MathAchSchool} data sets in the \pkg{nlme} package  [@PinheiroBates:2000]. \code{MathAchieve} comprises individual-level data on 7185  students in 160 high schools, and \code{MathAchSchool} contains school-level data:
```{r HSB-data}
data("MathAchieve", package="nlme")
dim(MathAchieve)
head(MathAchieve, 3)
tail(MathAchieve, 3)

@@

Some data-management is required before fitting a mixed-effects model to the HSB data:
```{r HSB-data-management, cache=TRUE}
HSB <- MathAchieve
HSB <- merge(MathAchSchool[, c("School", "Sector")],
             HSB[, c("School", "SES", "MathAch")], by="School")
names(HSB) <- tolower(names(HSB))
HSB <- within(HSB, {
  mean.ses <- ave(ses, school)
  cses <- ses - mean.ses
})
```
In the process, we merged variables from the school-level and student-level data sets and created two new school-level variables: \code{mean.ses}, which is the average SES for students in each school; and \code{cses}, which is each students' SES centered at their school means. For details, see @FoxWeisberg:2019 [Sec. 7.2.2].

Still following Fox and Weisberg, we proceed to use the \code{lmer()} function in the \pkg{lme4} package [@BatesEtAl:2015] to fit a mixed model for math achievement to the HSB data:
```{r HSB-lmer, cache=TRUE}
library("lme4", quietly=TRUE)
hsb.lmer <- lmer(mathach ~ mean.ses*cses + sector*cses
                   + (cses | school), data=HSB)
summary(hsb.lmer, correlation=FALSE)
```

@@

If the \code{clusterVariables} argument is omitted, then case-level CV is employed, with \code{k = 10} folds as the default, here each with $7185/10 \approx 719$ students. Notice that one of the 10 models refit with a fold removed failed to converge. Convergence problems are common in mixed-effects modeling. The issue here is that an estimated variance component is close to or equal to 0, which is at a boundary of the parameter space. That shouldn't disqualify the fitted model for the kind of prediction required for cross-validation.

\code{cv()} also has methods for mixed models fit by the \code{glmer()} function in the \pkg{lme4} package, the \code{lme()} function in the \pkg{nlme} package [@PinheiroBates:2000], and the \code{glmmTMB()} function in the \pkg{glmmTMB} package [@BrooksEtAl], along with a simple procedure for extending \code{cv()} to other classes of mixed-effects models. See the vignettes in the \pkg{cv} package for details.

@@

In this section, we introduce an artificial data set that exemplifies aspects of cross-validation particular to hierarchical models. Using this data set, we show that model comparisons employing cluster-based and those employing case-based cross-validation may not agree on a "best" model. Furthermore, commonly used measures of fit, such as mean-squared error, do not necessarily become smaller as models become larger, even when the models are nested, and even when the measure of fit is computed for the whole data set.

Consider a researcher studying improvement in a skill, singing, for example, among students enrolled in a four-year voice program at a music conservatory. The plan is to measure each student's skill level at the beginning of the program and every year thereafter until the end of the program, resulting in 5 annual measurements for each student. It turns out that singing appeals to students of all ages, and students enrolling in the program range in age from 20 to 70. Moreover, participants' untrained singing skill is similar at all ages, as is their rate of progress with training. All students complete the four-year program.
